---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Kendra Sultzer"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
library(ggplot2)
NTL.LTER <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
class(NTL.LTER$sampledate) #sampledate is a factor
NTL.LTER$sampledate <- as.Date(NTL.LTER$sampledate, format="%m/%d/%y")
class(NTL.LTER$sampledate) #now sampledate is a date


#2
mytheme <- theme_classic(base_size = 12) + 
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right", 
        legend.box.background = element_rect(color="black"), legend.background = element_blank(),  
        plot.title=element_text(hjust = 0.5))
#makes font size bigger, puts legend on right, puts box around legend, centers title in middle
theme_set(mytheme) #setting for all subsequent plots

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: slope (beta) of depth = 0
Ha: slope (beta) of depth is not equal to 0


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL.with.month <- mutate(NTL.LTER, month=month(sampledate))
NTL.subset <- NTL.with.month %>%
  filter(month==7)%>%
  select(lakename:daynum, depth:temperature_C)%>%
  na.omit(NTL.with.month)

#5
Temp.depth <- ggplot(data=NTL.subset, aes(x=depth, y=temperature_C))+
  geom_point()+
  geom_smooth(method = "lm")+
  ylim(0,35)+
  labs(title="Effect of Lake Depth on Temperature", x="Depth", y="Temperature (C)")
Temp.depth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Looking at the figure, we can see that as depth increases, temperature decreases. The distribution of these points suggests that the data might not be linear and more of the variance in the data might be captured by using a transformation. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
tempdepthlm <- lm(data=NTL.subset, temperature_C ~ depth)
summary(tempdepthlm)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Our model shows that there's a significant negative correlation between depth and temperature (lower temperatures at greater depths), and our model accounts for 73.87% of the total variance in temperature (p<0.05, df=9726). Since our p-value is significant, we can reject the null hypothesis and say that the beta of depth is not equal to zero. As depth increases by 1m, the temperature decreases by 1.946 degrees C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
Temp.AIC <- lm(data = NTL.subset, temperature_C ~ year4 + daynum + depth) 
step(Temp.AIC) #lowest AIC is all 3 variables

#10
#already defined model above but here are the results...
summary(Temp.AIC)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggested that we use the year, day of year, and depth to predict temperature in the multiple regression.  This updated model explains 74.11% of the variance in depth which was a slight improvement in variance over the model using only depth as the explanatory variable. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Temp.anova <- aov(data = NTL.subset, temperature_C ~ lakename) #Anova model
summary(Temp.anova)
Temp.anova2 <- lm(data = NTL.subset, temperature_C ~ lakename) #linear model
summary(Temp.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes there is significant difference in mean temperatures among the lakes (P<0.05).However, without post-processing, we can't tell which lakes are different from each other.     



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.width=9}
#14.
temp.depth.lake <- ggplot(data = NTL.subset, aes(x=depth, y=temperature_C, color=lakename))+
  geom_point(alpha=0.5)+
  geom_smooth(method="lm", se=FALSE)+
  labs(title="Effect of Depth on Temperature by Lakes", x = "Depth", y="Temperature (C)", color="Lake Name")+
  ylim(0,35)
temp.depth.lake

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Temp.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake and Paul Lake are the only 2 lakes that have the same mean temperature as Peter Lake (p>0.05).  No lakes have a mean temperature that is statistically distinct from all the other lakes.  Every lake's mean temperature is significantly similar to at least one other lake's mean temperature. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use a two-sample t-test to check if the mean temperature of Peter Lake is different than the mean temperature of Paul Lake. 


